### 2019-11-01 Writing/Palaeontology/draft.tex
Updated draft

\bigskip
\noindent RH: Characters correlation
\bigskip
\medskip
\begin{center}
\noindent{\Large \bf Influence of different modes of morphological character correlation on phylogenetic tree inference}
\bigskip

@@

\newpage
\begin{abstract}
Phylogenetic analysis algorithms assume character independence - a condition generally acknowledged to be violated by morphological data.
Correlation between characters can originate from intra-organismal features, shared phylogenetic history or forced by particular character-state coding schemes. 
Although the two first sources can be investigated by biologists \textit{a posteriori} and the third one can be diminished \textit{a priori} with careful character coding, phylogenetic software do not distinguish between any of them.
In this study, use a metric of raw character difference as a proxy for character correlation.
Using thorough simulations, we test the effect of increasing or decreasing character differences on tree topology.
Overall, we found a negative effect of reducing character correlations on recovering the correct topology.
%ap- will you confront the fact that this says that having character correlation is an advantage?
This means that dataset with low correlation between characters will make it more complicated to estimate a correct topology.
However, this effect is less important for matrices with a small number of taxa (25 in our simulations) where reducing or increasing character correlation is not more effective in getting the expected topology than randomly drawing characters.
Furthermore, in bigger matrices (350 characters), the inference method has a strong effect with Bayesian trees being consistently less affected by character correlation than maximum parsimony trees.
%ap but if cc increase recovering correct topology, then Bayesian is bad and we should use parsimony. Needs clarifying.
These results suggest that ignoring the problem of character correlation or independence can often impact topology in phylogenetic analysis.
However, encouragingly, they also suggest that, unless correlation is actively maximised or minimised, probabilistic methods can easily accommodate for a random correlation between characters.

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\section{Introduction}
The last two decades have witnessed a resurgence of interest in the use of morphological character data in phylogenetic studies.
This owes in large part to the use of fossils to undertake at least partial reconstructions of phylogenetic trees, especially where ancestral states reconstructions or absolute calibrations of divergence times are necessary. 
Although morphological character data are sometimes considered inferior to molecular sequence data due to the difficulties of collecting them and their sometimes subjectivity, they are usually the only source of phylogenetic data for extinct species.
While there is a general appreciation of the limits of morphological data, they are frequently dismissed without any empirical investigations into their statistical properties.
As morphological data are likely to continue to play an extensive role in phylogenetic analysis, it is essential to understand the circumstances under which morphological data might be expected to ``misbehave''.
This opens up possibilities for predicting problematic datasets and possibly proposing new confidence measures in phylogenetic datasets.
%MB: I think a light rewriting  of this opening paragraph would be good, focusing it more on a palaeontological audience (given the new target journal). We can cut the apologetics for morphology and focus it on palaeontology-relevant stuff  a bit more.

The non-independence of large numbers of morphological characters is often cited in anticipation of problems with morphological data \cite[e.g.][]{Davalos01072014, ZouConvergence}.
The assumption of character independence is central to phylogenetic inference methods such as maximum likelihood and maximum parsimony \citep[e.g.][]{joysey1982problems,felsenstein1985phylogenies,lewisa2001,felsenstein2004inferring}.
% This independence is assumed for both molecular and morphological characters \citep{huelsenbeck1999effect,Davalos01072014,ZouConvergence}.
However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differ in the size and number of states analysed.
 % (molecular matrices are generally larger with mainly four states - not including missing data or indels - and morphological matrices are smaller with generally two states but may have an arbitrary number of states (up to that supported by phylogenetic software; \citealt{Guillerme2016146}).
Second, correlations have been proposed to be stronger in morphological data
%(i.e. ``concentrated'' correlations)
but weaker and more ``spread-out'' in molecular data \citep[][]{huelsenbeck1999effect}. %i.e. ``diffused'' correlations - though some regions of the genome may be thought as more correlated than others
% Furthermore, because of the ``spread'' of correlations in molecular data, it can be harder to link the character correlation to a specific biological phenomenon (i.e. changes in morphological states can be more readily linked to a function \textit{cf.} changes in nucleotide states ; see below).
% Before discussing character correlation further,
Here we distinguish three different modes of forms of character correlation:
% MB: These considerations are interesting, and "modes" are also referred to in the title. The discussion does come back to these, but not sure it adds much. It could be an opportunity to remove almost a page and keep the overall flow more focused. Indeed, this is almost all said before (a bit more succinctly) by Mark Wilkinson.

\begin{itemize}
    \item \textbf{Intra-organismal dependence:} this is the result of an intrinsic biological link between two characters through development, pleiotropy, and/or biological function.
    For example the lower and upper molar characters in mammals generally occlude one another. 
    Therefore, one character describing a feature of a lower molar will be expected to be complemented by the surface of the occluding upper molar.
    Characters of the occlusal surface of two opposing molars will be expected to directly covary.
    % Pleiotropy also results in covariation between different aspects of phenotype and may thus induce correlations. 
    % From a phylogenetic perspective, it can be especially pernicious because the relationship between the traits in question may have no obvious link from a morphological or functional comparison alone. 
    Intra-organismal links can be the targets of comparative developmental biology \citep{goswami2010,kelly2010reduced,stoessel2013morphological,goswami2014macroevolutionary} or functional investigations. %AP - I would drop this sentence and say whether this type of correlation is useful or not in less uncertain terms.

    \item \textbf{Evolutionary dependence:} this is the result of sets of characters co-evolving due to selection, likely related to functional links between two traits that help serve an overall lifestyle trait \citep{ClarkeMiddleton2008}. 
    Unlike the case of intra-organismal dependence, there need not be an intrinsic constraint that causes these traits to covary.
    For example, in vertebrates, axial elongation can be correlated to limb reduction with snake-like bodies evolving multiple times in numerous tetrapod lineages.
    % This is thought to correspond to adaptations for fossoriality or aquatic lifestyles.
    % Such covariances are generally studied in the context of a given phylogeny, often one derived from molecular data with the morphological traits of interest mapped on it.
    Many methods have been developed to study these correlations, especially since they can provide us with a lot of information on how specific groups acquired specific characteristics \citep{Lande1983,Maddison1990,Pagel1994,Pagel2006,Grabowski2016}.
    However, these methods do not give us a means to objectively control correlations that might adversely affect phylogenetic inference. %AP this is good.

    \item \textbf{Coding dependence:} this is the results of researcher methodology for defining or/and coding discrete morphological characters \citep{Brazeau2011,simoes2017giant}.
    Coding dependence manifests itself in several ways, particularly in coding redundant information.
    For instance, coding for the same absence in different characters creates state transformations associated with the loss or gain of a particular character.
    This occurs when a number of multistate characters include two variable feature states (e.g. large, small; red, blue etc.) in conjunction with absence.
    For example, two characters ``tail colour'' and ``tail length'' could be coded two times as an absence for a taxon with no tail.
    % It is worth noting, however, that these correlations could also be due to the nature of the available data, especially in palaeontology.
    % For example, when only one fragmentary molar is available to describe a specimen, researchers have to ``extract'' as much phylogenetic information from the available data as possible, potentially inducing correlations.
    This coding dependency is linked to hierarchical dependency between characters \citep{wilkinson1995coping,BrazeauNA}. %AP-Can cut this sentence
    % In the example above, attributes of the tail (colour or length) are hierarchically dependent on the presence of the tail in the first place.
    % Finally this can also be due to a bias in the amount of characters available.
    % For example, in skulls, because of their complexity, there is a high likelihood of inducing correlation (by effectively reducing structural complexity to discrete characters).
\end{itemize}

\noindent Of course, the three sources of dependence can have an interaction: characters describing the left and right lower/upper molars will have induced dependence due to the modularity of the molars, their shared history and the duplicated coding.
Logical dependence, however, is easily distinguished prior to phylogenetic inference, while intra-organismal and evolutionary dependence are much harder to parse. %AP-what is 'logical dependece'?  its not defined... I guess you meaning "coding'... why is it easy to distinguish?
Meanwhile, the development of algorithms and software has not yet caught up with the need to deal with these interdependencies \citep{de2015parsimony,BrazeauNA}.
Intra-organismal dependence requires more detailed, often extremely time-consuming studies (possibly beyond the limits of available technology).
% Even after all of the effort is expended, the results might then only be known for a single (model) species.
Evolutionary dependence itself requires the accurate resolution of a phylogenetic tree, and is best determined by independent character sets. 
% This is frequently accomplished by mapping morphological traits on molecular phylogenetic trees.

% These sources of dependence between characters are well studied in biology.
Biological and evolutionary dependences are inherent parts to Evo-Devo and macroevolutionary studies and best practices to avoid coding-induced dependences are commonly known and applied.
However, eventually, all these characters, whether they are independent or not are analysed through phylogenetic inference software that are blind to these distinctions.
If fact, what the software are confronted with is a two dimensional matrix problem that renders all morphological dependencies invisible.
This introduces a new, less studied, source of character dependence:

\noindent \textbf{Characters compatibility}: this is the result of how software actually interprets the differences between characters.
% MB: Well, software can maximise it, but compatibility itself is just a property of the data.
The majority of phylogenetic algorithms cannot really interpret character descriptions and state meanings (i.e. character ontologies), and instead must interpret discrete tokens as signifiers of shared similarity.
Therefore, phylogenetic character matrices that contain a great number of characters but a few number of tokens can easily lead to correlation between characters. %AP I do not understand this definition.  The example did not clear it up.  What are tokens?
% MB: I am having trouble interpreting this.
For example, if we consider the following matrix containing four cetartiodactyls - say a pig (\textit{Sus}), a deer (\textit{Cervus}), a hippo (\textit{Hippopotamus}) and a whale (\textit{Balaenoptera}) - and four binary characters - say (\textbf{C1}: presence (1) or absence (0) of an astragalus; \textbf{C2}: presence (0) or absence (1) of baleen; \textbf{C3}: presence (0) or absence (1) of a left astragalus with a double pulley; \textbf{C4}: presence (0) or absence (1) of a right astragalus with a double pulley.

\begin{table}
\center
    \begin{tabular}{r|cccc}
            & C1 & C2 & C3 & C4\\
        \hline
        \textit{Sus} & 1 & 1 & 1 & 1\\
        \textit{Cervus} & 1 & 1 & 1 & 1\\
        \textit{Hippopotamus} & 1 & 1 & 1 & 1\\
        \textit{Balaenoptera} & 0 & 0 & 0 & 0\\
    \end{tabular}
    \caption{Example of a matrix with software induced character correlation. \textbf{C1}: presence (1) or absence (0) of an astragalus (ankle bone); \textbf{C2}: presence (0) or absence (1) of baleens; \textbf{C3}: presence (0) or absence (1) of a left astragalus with a double pulley; \textbf{C4}: presence (0) or absence (1) of a right astragal with a double pulley.}
    \label{Tab:example_matrix}
\end{table}

In the example in Table \ref{Tab:example_matrix}, the characters \textbf{C1} and \textbf{C2} are the most likely to be truly independent; characters \textbf{C3} and \textbf{C4} suffer from a a coding induced dependency; characters \textbf{C1} and \textbf{C3}/\textbf{C4} have an evolutionary induce dependency and again, characters \textbf{C2} and \textbf{C3}/\textbf{C4} are likely to be independent.
Yet a phylogenetic software will treat all these four characters in exactly the same way.
% In fact, for practical reasons, phylogenetic software do not (or rarely) distinguish between the different sources of correlation.
In this specific example, we would intuitively want to favour character \textbf{C1} and \textbf{C2} that seem to be more likely to reflect the split by separating artiodactyles from the cetaceans.
% In other words, this is the correlation between characters that depicts the evolutionary history of this group.
However, the other characters depict the same history %(i.e. the same split)
 but for the wrong reasons. % (note that even if \textbf{C2} was coded presence (1) or absence (0) of baleen, the same split would still be estimated by the software).
One solution would be to a) weight some characters (e.g. giving \textbf{C3} and \textbf{C4} a lower weight) and b) make some characters dependent on each other (e.g. making \textbf{C3} and \textbf{C4} dependent on \textbf{C1}).
However, this can also be a bad practice since it can easily lead to the resulting tree reflecting the researcher's prior expectations rather than the true evolutionary history.
Alternatively, partitioning strategies have been used to reduce this problem \citep[see][and references within]{Rosa2019}.
This example illustrates how some characters will therefore be expected to covary in a non-phylogenetic way, and that this phenomenon can reasonably be expected to mislead phylogenetic analysis. %AP-what is "covary in a non-phlyogenetic way'??  this section-I don't get it.
% MB: Is all of this necessary?

How do these specific correlation really affect topology?
Although it has been tackled empirically (for morphological data \citealt{Davalos01072014}, and molecular data \citealt{ZouConvergence}), the question has never been explored through a thorough simulation framework. %AP I hope we come back to these studies in the discussion.
% MB: You may wish to add one line to say what the fundamental importance of doing so is. What will this paper add that those did not?
% In the specific case of discrete morphological characters, how does the reduc phenotypic information and its different sources of correlation is 
% More specifically, how can phylogenetic inference software distinguish between the complex sources of correlation tha
% In other words, how can we distinguish between the complex sources of correlation in phenotype using only discrete morphological characters?
% \cite{parins2017use}
Here we formally assess the effect of discrete characters' correlation using simulated data.
We propose % a new
using the Hamming distance metric to measure the difference between characters (as a proxy for these three sources of correlation as interpreted by the software) and a protocol to modify discrete morphological matrices to increase or decrease the overall differences or similarities between characters.
We found that overall, there is a detectable effect of character correlation on topology where a decrease in character dependence results in a decrease in the ability to recover the correct topology.
These results, however, vary greatly in magnitude depending on the size of matrix and the inference method used.

\section{Methods}
To assess the effects of character correlation on the accuracy of phylogenetic inference we generated a series of matrices exhibiting different levels of correlation between some characters (Fig.\ref{Fig:outline} - note that each step is described in more details below): %AP different average correlation between characters?
\begin{enumerate}
    \item \textbf{Simulating matrices}: we simulated discrete morphological matrices with 25, 75 or 150 taxa, for 100, 350 or 1000 characters, hereafter called the ``normal'' matrices. This step resulted in 9 matrices.

@@

\subsection{Measuring differences between characters}
\label{CDdescription}
To measure the effect of character correlation as interpreted by the phylogenetic software, we define characters as being entirely correlated if they give the same phylogenetic information.
%(i.e. entirely correlated characters would define identical splits between a set of taxa).
In order to measure this, we used the following scaled Hamming distance between two characters:

\subsubsection{Character Difference ($CD$)}
\begin{equation}
    CD_{(x,y)} = \frac{\sum_{i}^{n}\abs{x_{i} - y_{i}}}{n-1}
\end{equation}
%CD is similar to https://onlinelibrary.wiley.com/doi/full/10.1111/pala.12384 but by character rather than by species

\noindent Where $n$ is the number of taxa with comparable characters $x$, $y$ and $x_i$, $y_i$ are each character's state for the $i^{th}$ taxon.
$CD$ is a continuous scaled Hamming distance metric bounded between $0$ and $1$.
Since we are considering differences as being only Fitch-like (non-additive) and unweighted, we calculated the difference between character states as binary %.
% Two same character states tokens have a difference of zero and two different ones have a difference of one
(e.g. $2 - 2 = 0$ or $1 - 8 = 1$).
% Additionally, we only consider differences for taxa with shared information \citep{GowerDist}.
% For example, for two characters X and Y for four taxa, if character \texttt{X = \{0,0,?,1\}} for the first, second, third and fourth taxa respectively and character \texttt{Y = \{0,0,1,1\}}, only the first, second and fourth taxa are compared for this character.
% Note that in these simulations, no missing data was included so the characters are compared for equal number of taxa (see below).

We standardised each character by arbitrarily modifying their character state tokens (or symbols) by order of appearance.
We replaced all the occurrences of the first token to be $1$, the second to be $2$, etc.
This procedure allows comparison of characters regardless of the significance given to their tokens (following the \textit{xyz} notation in \citealt{felsenstein2004inferring}; as used in \citealt{Davalos01072014}).
This way, a character \texttt{A = \{2,2,3,0,0,3\}} for six taxa would be standardised as \texttt{A' = \{1,1,2,3,3,2\}}.
% Note that in terms of phylogenetic signal, both \texttt{A} and \texttt{A'} are exactly identical (forming three distinct splits in the tree inference process).
When the character difference is null ($0$) it means that characters convey the same phylogenetic signal (i.e. characters are entirely correlated). 
When the character difference is maximal ($1$) it means the characters convey the greatest difference in phylogenetic signal. % (i.e. characters are uncorrelated).
%However, a character difference of $0$ %(i.e. the same phylogenetic signal)
%does not mean the opposite of $1$ (i.e. \textit{not} the opposite phylogenetic signal but the most different number of implied splits) . %AP this sentence was awkward or ungrammatical or incorrect, not sure which.  Don't talk about "opposites" it doesnt mean anything.
% For example with three characters \texttt{A = \{0,1,2,1\}}, \texttt{B = \{2,0,1,0\}} and \texttt{C = \{0,0,1,2\}}, $CD_{(A,B)} = 0$ and $CD_{(A,C)} = 1$ (translated into \texttt{A' = \{1,2,3,2\}} \texttt{B' = \{1,2,3,2\}} and \texttt{C' = \{1,1,2,3\}}; $CD_{(A,B)} = (0+0+0+0)/(4-1) = 0$ and $CD_{(A,C)} = (0+1+1+1)/(4-1) = 1$).
% Because the character is continuous and bounded between $(0,1)$, it can be interpreted as the probability of two characters leading to a different set of splits (i.e. a different phylogenetic signal).

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146} using \texttt{R} \citep{R}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution with $\lambda$ $>$ $\mu$\citep[\texttt{diversitree} v0.9-8;][]{fitzjohndiversitree2012} and saving the trees after reaching either 25, 75 or 150 taxa (see supplementary material 3, Figs 1, 2 and 3).
For each tree, we arbitrarily set the outgroup to be the first taxon (alphabetically) thus effectively rooting the trees on this taxon.
These trees are hereafter called the ``true'' trees (see distinction below).
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The ``morphological HKY-binary'' model \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a Dirichlet distribution $Dir(1,1,1,1)$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    % This model has the advantage of not favouring Bayesian inference \citep[; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We drew the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep{Guillerme2016146,ZouConvergence}.
    % This model assumes a equal transition rate between character states which might seem overly simplistic, excluding other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    % However, \cite{Wright01072016} have shown that equal transition rate is a reasonable assumption for many empirical datasets.
\end{itemize}

\noindent For each character, both models (``morphological HKY-binary'' or M$k$) were chosen randomly and run with an overall evolutionary rate drawn from a gamma distribution ($\beta$ = $100$ and $\alpha$ = $5$).
These low evolutionary rate values allowed reduction in the number of homoplasic character changes, thus reinforcing the phylogenetic information in the matrices.
We re-simulated every invariant characters to obtain a matrix with no invariant characters in order to better approximate real morphological data matrices.
To ensure that our simulations were reflecting realistic observed parameters, we only selected matrices with Consistency Indices (CI) superior to $0.26$ \citep{sanderson1989patterns,OReilly20160081}.
% Note that both models (``morphological HKY-binary' and Mk) are both based on a Markovian model and only differ in the number of states generated and transformed.
% However, the ``morphological HKY model'' is not the one \textit{directly} estimated in the phylogenetic software used in this protocol (MrBayes, see below).

For each tree with 25, 75 or 150 taxa we generated matrices with 100, 350 and 1000 characters following \cite{OReilly20160081}.
The matrices were generated using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
% To estimate the variance of our simulations and assess the effect of our random parameters,
We repeated this step 20 times resulting in 180 ``normal'' morphological matrices.

\subsection{Modifying the matrices}
We calculated the pairwise character differences for each generated matrix using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We then modified the matrices to either maximise or minimise the pairwise character differences for each matrix using three different algorithms.
For maximising the pairwise differences between characters, we selected the characters that were the most similar to all the others (mean character difference $<0.25$) and replaced them randomly by any of the remaining characters.
This operation increased the overall pairwise character difference in the matrix thus making the characters more dissimilar.
Conversely, for minimising the pairwise character differences, we selected the most dissimilar characters (mean character difference $>0.75$) and randomly replaced them with the remaining ones.
Finally, because this operation effectively changes the weight of characters (i.e. giving the characters $<0.25$ or $>0.75$ a weight of $0$ and giving the randomly selected remaining characters a weight of +$1$), we randomly replaced the average number of characters replaced in the character maximisation and minimisation by any other characters as a randomised expectation modification.% (i.e. randomly weighting characters).
Each replaced characters were randomly selected by sampling them randomly from the list of the remaining characters.
% Each of the three matrices are effectively a bootstrap pseudo-replication of the ``normal'' matrix with the ``randomised'' one being a random one and the ``maximised'' and ``minimised'' being conditional bootstraps.
This step resulted in a total of 720 matrices (hereafter called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices - see Fig. \ref{Fig:modif_matrix} for an illustration).
The algorithms for the three modifications are available on GitHub (\url{https://github.com/TGuillerme/CharactersCorrelation})

@@

\subsection{Inferring topologies}
We inferred the topologies with both Bayesian inference and maximum parsimony using MrBayes \citep[v3.2.6;][]{Ronquist2012mrbayes} and PAUP* \citep[v4.0a151;][]{swofford2001paup} respectively.
For both methods, we used the arbitrarily chosen outgroup in the simulations to root our trees (the first taxon in the taxa list).
The maximum parsimony inference was run using a heuristic search with random sequence addition replicate 100 times with a limit of $5\times10^6$ rearrangements per replicates.
 % (\texttt{hsearch addseq=random nreps=100 rearrlimit=5000000 limitperrep=yes}).

>>>>

\bigskip
\noindent RH: Characters correlation
\bigskip
\medskip
\begin{center}
\noindent{\Large \bf Influence of morphological character correlation on phylogenetic tree inference}
\bigskip

@@

\newpage
\begin{abstract}
Phylogenetic analysis algorithms assume character independence - a condition generally acknowledged to be violated by morphological data.
Correlation between characters can originate from intra-organismal features, shared phylogenetic history or forced by particular character-state coding schemes. 
Although the two first sources can be investigated by biologists \textit{a posteriori} and the third one can be diminished \textit{a priori} with careful character coding, phylogenetic software do not distinguish between any of them.
In this study, use a metric of raw character difference as a proxy for character correlation.
Using thorough simulations, we test the effect of increasing or decreasing character differences on tree topology.
Overall, we found a negative effect of reducing character correlations on recovering the correct topology.
This means that dataset with low correlation between characters will make it more complicated to estimate a correct topology.
However, this effect is less important for matrices with a small number of taxa (25 in our simulations) where reducing or increasing character correlation is not more effective in getting the expected topology than randomly drawing characters.
Furthermore, in bigger matrices (350 characters), the inference method has a strong effect with Bayesian trees being consistently less affected by character correlation than maximum parsimony trees.
These results suggest that ignoring the problem of character correlation or independence can often impact topology in phylogenetic analysis.
However, encouragingly, they also suggest that, unless correlation is actively maximised or minimised, probabilistic methods can easily accommodate for a random correlation between characters.

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\section{Introduction}
The last two decades have witnessed a resurgence of interest in the use of morphological character data in phylogenetic studies.
This owes in large part to the advent of new models, methods and software \citep{lewisa2001,ronquista2012,Ronquist2012mrbayes,heath2014fossilized} and the use of fossils to undertake at least partial reconstructions of phylogenetic trees, especially where ancestral states reconstructions or absolute calibrations of divergence times are necessary.
While there is a general appreciation of the limits of morphological data, they are frequently dismissed without any investigations into their statistical properties.
As morphological data are likely to continue to play an extensive role in phylogenetic analysis (namely for neontological analysis), it is essential to understand the circumstances under which morphological data might be expected to lead to erroneous phylogenetic inferences.
This opens up possibilities for analysing problematic datasets and possibly proposing new confidence measures in phylogenetic datasets.

The non-independence of large numbers of morphological characters is often cited in anticipation of problems with morphological data \cite[e.g.][]{Davalos01072014, ZouConvergence}.
The assumption of character independence is central to phylogenetic inference methods such as maximum likelihood and maximum parsimony \citep[e.g.][]{joysey1982problems,felsenstein1985phylogenies,lewisa2001,felsenstein2004inferring}.
However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differ in the size and number of states analysed.
Second, correlations have been proposed to be stronger in morphological data but weaker and more ``spread-out'' in molecular data \citep[][]{huelsenbeck1999effect}.
Here we distinguish three different modes of forms of character correlation (described in more details in \citealt{wilkison1992ordered,wilkinson1995character,wilkinson1995coping}):

\begin{itemize}
    \item \textbf{Intra-organismal dependence:} this is the result of an intrinsic biological link between two characters.
    For example the occlusal surface of two opposing molars will be expected to directly covary.

    \item \textbf{Evolutionary dependence:} this is the result of sets of characters co-evolving. 
    For example, in vertebrates, axial elongation can be correlated to limb reduction with snake-like bodies evolving multiple times in numerous tetrapod lineages.

    \item \textbf{Coding dependence:} this is the results of researcher methodology for defining or/and coding discrete morphological characters \citep{Brazeau2011,simoes2017giant}.
    For example, two characters ``tail colour'' and ``tail length'' could be coded two times as an absence for a taxon with no tail \citep{wilkinson1995coping,BrazeauNA}.
\end{itemize}

\noindent Of course, the three sources of dependence can have an interaction: characters describing the left and right lower/upper molars will have induced dependence due to the modularity of the molars, their shared history and the duplicated coding.
Coding dependence, however, is easily distinguished prior to phylogenetic inference, while intra-organismal and evolutionary dependence are much harder to parse.
Meanwhile, the development of algorithms and software has not yet caught up with the need to deal with these interdependencies \citep{de2015parsimony,BrazeauNA}.
Intra-organismal dependence requires more detailed, often extremely time-consuming studies (possibly beyond the limits of available technology).
Evolutionary dependence itself requires the accurate resolution of a phylogenetic tree, and is best determined by independent character sets.

Biological and evolutionary dependences are inherent parts to Evo-Devo and macroevolutionary studies and best practices to avoid coding-induced dependences are commonly known and applied.
However, eventually, all these characters, whether they are independent or not are analysed through phylogenetic inference software that are blind to these distinctions.
If fact, what the software are confronted with is a two dimensional matrix problem that renders all morphological dependencies invisible.
This introduces a new, less studied, source of character dependence:

\begin{itemize}
    \item \textbf{Characters compatibility}: this is the result of how software actually interprets the differences between characters.
    The majority of phylogenetic algorithms cannot interpret character descriptions and state meanings, and instead must interpret discrete states as signifiers of shared similarity: for each characters, species are grouped if they have identical characters.
    Therefore, character with the same number of states will have a chance of being correlated regardless of mode of form of character correlation.
\end{itemize}

\noindent How does character correlation really affect topology?
Although it has been tackled empirically (for morphological data \citealt{Davalos01072014}, and molecular data \citealt{ZouConvergence}), the question has never been explored through a thorough simulation framework.
Simulations can allow the design of dataset with controlled levels of correlation and for which the correct expected answer is known by design.
Furthermore, using simulations will allow to draw generalities on the effect of character correlation on phylogenetic inference conversely to potential idiosyncrasies of empirical datasets.
Here we use a Hamming distance metric to measure the difference between characters (as a proxy for these three sources of correlation as interpreted by the software) and a protocol to modify discrete morphological matrices to increase or decrease the overall differences or similarities between characters.
We found that overall, there is a clear effect of character correlation on topology where a decrease in character correlation results in a decrease in the ability to recover the correct topology.
These results, however, vary greatly in magnitude depending on the size of matrix and the inference method used.

\section{Methods}
To assess the effects of character correlation on the accuracy of phylogenetic inference we generated a series of matrices exhibiting different average correlation between characters (Fig.\ref{Fig:outline} - note that each step is described in more details below):
\begin{enumerate}
    \item \textbf{Simulating matrices}: we simulated discrete morphological matrices with 25, 75 or 150 taxa, for 100, 350 or 1000 characters, hereafter called the ``normal'' matrices. This step resulted in 9 matrices.

@@

\subsection{Measuring differences between characters}
\label{CDdescription}
To measure the effect of character correlation as interpreted by the phylogenetic software, we define characters as being entirely correlated if they give the same phylogenetic information.
In order to measure this, we used the following scaled Hamming distance between two characters:

\subsubsection{Character Difference ($CD$)}
\begin{equation}
    CD_{(x,y)} = \frac{\sum_{i}^{n}\abs{x_{i} - y_{i}}}{n-1}
\end{equation}

\noindent Where $n$ is the number of taxa with comparable characters $x$, $y$ and $x_i$, $y_i$ are each character's state for the $i^{th}$ taxon.
$CD$ is a continuous scaled Hamming distance metric bounded between $0$ and $1$.
Since we are considering differences as being only Fitch-like (non-additive) and unweighted, we calculated the difference between character states as binary (e.g. $2 - 2 = 0$ or $1 - 8 = 1$).

We standardised each character by arbitrarily modifying their character state tokens (or symbols) by order of appearance.
We replaced all the occurrences of the first token to be $1$, the second to be $2$, etc.
This procedure allows comparison of characters regardless of the significance given to their tokens (following the \textit{xyz} notation in \citealt{felsenstein2004inferring}; as used in \citealt{Davalos01072014}).
This way, a character \texttt{A = \{2,2,3,0,0,3\}} for six taxa would be standardised as \texttt{A' = \{1,1,2,3,3,2\}}.
When the character difference is null ($0$) it means that characters convey the same phylogenetic signal (i.e. characters are entirely correlated). 
When the character difference is maximal ($1$) it means the characters convey the greatest difference in phylogenetic signal

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146} using \texttt{R} \citep{R}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution with $\lambda$ $>$ $\mu$\citep[\texttt{diversitree} v0.9-8;][]{fitzjohndiversitree2012} and saving the trees after reaching either 25, 75 or 150 taxa (see supplementary material 3, Figs 1, 2 and 3).
For each tree, we arbitrarily set the outgroup to be the first taxon (alphabetically) thus effectively rooting the trees on this taxon.
These trees are hereafter called the ``true'' trees (see distinction below).
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The ``morphological HKY-binary'' model \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a Dirichlet distribution $Dir(1,1,1,1)$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We drew the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep{Guillerme2016146,ZouConvergence}.
\end{itemize}

\noindent For each character, both models (``morphological HKY-binary'' or M$k$) were chosen randomly and run with an overall evolutionary rate drawn from a gamma distribution ($\beta$ = $100$ and $\alpha$ = $5$).
These low evolutionary rate values allowed reduction in the number of homoplasic character changes, thus reinforcing the phylogenetic information in the matrices.
We re-simulated every invariant characters to obtain a matrix with no invariant characters in order to better approximate real morphological data matrices.
To ensure that our simulations were reflecting realistic observed parameters, we only selected matrices with Consistency Indices (CI) superior to $0.26$ \citep{sanderson1989patterns,OReilly20160081}.

For each tree with 25, 75 or 150 taxa we generated matrices with 100, 350 and 1000 characters following \cite{OReilly20160081}.
The matrices were generated using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We repeated this step 20 times resulting in 180 ``normal'' morphological matrices.

\subsection{Modifying the matrices}
We calculated the pairwise character differences for each generated matrix using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We then modified the matrices to either maximise or minimise the pairwise character differences for each matrix using three different algorithms.
For maximising the pairwise differences between characters, we selected the characters that were the most similar to all the others (mean character difference $<0.25$) and replaced them randomly by any of the remaining characters.
This operation increased the overall pairwise character difference in the matrix thus making the characters more dissimilar.
Conversely, for minimising the pairwise character differences, we selected the most dissimilar characters (mean character difference $>0.75$) and randomly replaced them with the remaining ones.
Finally, because this operation effectively changes the weight of characters (i.e. giving the characters $<0.25$ or $>0.75$ a weight of $0$ and giving the randomly selected remaining characters a weight of +$1$), we randomly replaced the average number of characters replaced in the character maximisation and minimisation by any other characters as a randomised expectation modification.
Each replaced characters were randomly selected by sampling them randomly from the list of the remaining characters.
This step resulted in a total of 720 matrices (hereafter called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices - see Fig. \ref{Fig:modif_matrix} for an illustration).
The algorithms for the three modifications are available on GitHub (\url{https://github.com/TGuillerme/CharactersCorrelation})

@@

\subsection{Inferring topologies}
We inferred the topologies with both Bayesian inference and maximum parsimony using MrBayes \citep[v3.2.6;][]{Ronquist2012mrbayes} and PAUP* \citep[v4.0a151;][]{swofford2001paup} respectively.
For both methods, we used the arbitrarily chosen outgroup in the simulations to root our trees (the first taxon in the taxa list).
The maximum parsimony inference was run using a heuristic search with random sequence addition replicate 100 times with a limit of $5\times10^6$ rearrangements per replicates.

Bayesian inference was run using an M\textit{k} model with ascertainment bias and four discrete gamma rate categories (M\textit{kv} $4\Gamma$) with an variable rate prior an exponential (0.5) shape.
The MCMC was ran over two runs of 6 chains each for a maximum of $1\times10^9$ generations with a sampling every 200 generations with an automatic stop if the average standard deviation of split frequency (ASDSF) fell below 0.01.
Due to cluster hardware requirements and to save some time, when chains didn't converge and the runs exceeded 5GB each, we aborted the MCMC and computed the consensus tree from the unconverged chains.
In practice, these few MCMC got stuck at an ASDSF around (but not below) 0.01.
